\fontfamily{phv}\selectfont{Causal Discovery methods for Effective Connectivity in Human Brains}

### 2014-01-17 docs/paper/paper final.tex
margins weggehaald, scheelt een halve pagina
Fixed bibliography typos. Addition on unfaithfulness check.

% Margins & Headers
\marginsize{2.5cm}{2.5cm}{3.0cm}{2.0cm}
\columnsep 0.4in
\footskip 0.4in 
\usepackage{changepage}

@@

\paragraph{Run-time performance}
The PC algorithm run-time is in the order of the amount of dependency tests to be performed and thus it has a worst-case runtime complexity exponential in the maximal branching degree present in the graph.
However, this worst-case scenario only happens when no edges can be removed; all subsets of the complete set of nodes have to be used to test d-separation of every pair of nodes.
For a sparse graph, many edges may be removed for small separating sets, and this enables the standard PC algorithm to often run fast in practice perform well on hundreds of variables.

@@

%EMS-PC extension
\paragraph{EMS-PC}
We have combined the above ideas - multiple separating sets, unfaithfulness check and explicit orientation check - into an adaptation of PC, which we refer to as Explicit Multiple Separating Set PC (EMS-PC).
These three changes have been made in this adaption with respect to standard PC as described in algorithm~\ref{alg:pc_alg}.
At line~\ref{alg:pc_alg:next_pair}, the search for separating sets for $X$ and $Y$ stops when a separating set is found; instead, EMS continues the search for more separating sets of the same order.
All of these separating sets are recorded.

@@

\subsection{Experimental validation}
\paragraph{Experimental pipeline}\label{sec:pipeline}
% What type of data is needed
PC algorithm requires observational data where each instance is a vector of single observations per variable.
We can model functional resting-state time-series as coming from some stable underlying mechanism that is largely time-independent.
Resting-state data can be intuitively thought of being more time independent than task-influenced data due to the emergent character of activity pattens that occur when in resting-state.
Nevertheless it is most likely not a fully correct modelling assumption, though enables straightforward application of standard and EMS-PC to functional data without the problem of HRF.

% Application to data
We applied standard PC and EMS-PC on functional resting-state time-series data of six healthy individuals, also previously used by Hinne et al.\cite{hinne2013}.
1030 volumes were obtained per subject.
After preprocessing the fMRI data was parcellated according to the Automated Anatomical Labeling (AAL) atlas in 116 regions of interest.
The exact procedure of subject acquisition, used scanner properties and parameters and data preprocessing is provided in appendix~\ref{sec:data}.
We use Fisher's z-test for conditional independence to match the continuous nature of the data.

% Implementation
We have implemented standard PC and EMS-PC in Matlab using statistical tests and ideas from the Bayesian Network Toolbox for Matlab.\footnote{Bayesian Network Toolbox for Matlab: https://github.com/bayesnet/bnt}
This toolbox also includes an implementation for standard PC, however to gain better insight in the workings and to have greater flexibility in application, we have implemented standard PC separately.
The Matlab scripts used for the results are made freely available online.\footnote{Matlab implementation of EMS-PC: https://github.com/tomderuijter/EMS-PC}

@@

% Parameter selection
Due to the lack of a baseline to compare results with, dependency thresholds were chosen by hand.
All experiments listed here were run with an alpha value of 0.01 for the independence test.
The number of random permutations was set on 1000 per subject, which was found to be enough to determine statistical significance for 99\% of the found edges with a binomial statistical test.

@@

\section{Data acquisition}\label{sec:data}
%Subject acquisition and instruction
%Used equipment and specs
%Experimental setup
%Scanner settings
%Preprocessing
%- Software
%- Scripts
%Resulting data specs
We performed our analysis on the same data as treated by Hinne et al. in the aforementioned Bayesian structural analysis \cite{hinne2013, hinne2013structfunc}.
This article describes how resting-state functional data of six subjects was measured at the Donders Centre for Cognitive Neuroimaging at the Radboud University.
This functional data is divided in 116 nodes and 1029 time-series per subject.
Additionally, diffusion-weighted images of these subjects have been obtained.
Based on these DWI-measurements, a structural connectivity matrix has been inferred with a technique described by Hinne et al.
We used the six resulting resting-state functional time series data and connectivity matrices from this Bayesian analysis.

All measurements and pre-processing as described in this section has been done during previous research. No additional pre-processing was needed in our research in order to apply the PC algorithm on this data.

\section{Results}\label{sec:results}
Due to the size of the resulting graphs, we represent them in the form of connectivity matrices for convenience.
A connection from region $n$ to region $m$ is represented by point $(n,m)$ in the matrix.
Higher values represent higher frequencies of a connection over all permutations.
In all matrices, indices 1-45 represent the brain regions of the left hemisphere, indices 45-90 represent the right hemisphere, and indices 91-116 represent regions in the cerebellum.
As such, the matrices are divided into blocks: the top left corner represents the connections within the left hemisphere, the bottom left corner represents the connections between the left hemisphere and the cerebellum, etc.

\subsection{Structural graphs}
% Measuring whole-brain structural and directional consistency of PC - frequency of occurrences within subjects
% Applying the standard PC algorithm yields an undirected graph, in which each edge has a certainty.
% This is done both per subject, and as an avarage over all subjects.
% We compared the generated structure with a Bayesian inference approach on finding structural connectivity from diffusion weighted imaging, as described in Hinne et al.\cite{hinne2013}.
Modified PC and EMS-PC do not differ in finding the structure.
As edges found with the structural part of PC are undirected, the found matrices are symmetrical.
The results are presented per subject in figure~\ref{fig:struct_subjects} and as an average across all subjects in figure~\ref{fig:struct_avg}.
For comparison, the structural connectivity as found by Hinne et al.\cite{hinne2013} are also show.

@@

\subsection{Directed graphs}
This structural graph has been used as a basis for a directed graph, for both standard PC and EMS-PC.
As there were no notable differences between different subjects, only the average PDAG across all subjects is presented.
The results for standard PC and EMS-PC are shown in figure~\ref{fig:pdag_avg}.
Although many arrowheads are found with probabilities above 0.9 for both variants of the algorithm, the graph looks strongly symmetric.
In terms of causality, this means that many latent sources are still present and causal relations between the measured brain regions themselves are not clearly visible.

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.49\textwidth]{images/new/PDAG_avg_mod-crop}
  \includegraphics[width=0.49\textwidth]{images/new/PDAG_avg_expl-crop}
  \caption{The PDAG found standard PC at the left, and EMS-PC at the right. A high value denotes a high certainty of an arrowhead. The average of all six subjects is taken.}
  \label{fig:pdag_avg}
\end{figure}

Another representation of this data has been used to emphasise the asymmetric parts of the directed graphs.
To achieve this, the absolute difference between these matrices and their transpose has been calculated, lifting out those points for which there is an arrow in one direction, but not in the opposite direction.
These resulting antisymmetric matrices are presented per subject for EMS-PC in figure~\ref{fig:asym_subjects}. The average across all subjects for both variants of the PC algorithm are presented in figure~\ref{fig:asym_avg}.

\begin{figure}[h!]
  \centering
  \includegraphics[height=0.25\textwidth]{images/new/assym_subj1_expl-crop}
  \includegraphics[height=0.25\textwidth]{images/new/assym_subj2_expl-crop}
  \includegraphics[height=0.25\textwidth]{images/new/assym_subj3_expl-crop}
  \includegraphics[height=0.25\textwidth]{images/new/assym_subj4_expl-crop}
  \includegraphics[height=0.25\textwidth]{images/new/assym_subj5_expl-crop}
  \includegraphics[height=0.25\textwidth]{images/new/assym_subj6_expl-crop}
  \caption{The assymetric parts of the PDAG found with EMS-PC for all six subjects, in which a high value denotes a high certainty of a one-directional arrow.}
\label{fig:asym_subjects}
\end{figure}

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.49\textwidth]{images/new/assym_avg_mod-crop}
  \includegraphics[width=0.49\textwidth]{images/new/assym_avg_expl-crop}
  \caption{The asymmetric parts of the PDAG found with standard PC on the left, and EMS-PC on the right. A high value denotes a high certainty of an one-directional arrow. The average of all six subjects is taken.}
  \label{fig:asym_avg}
\end{figure}

@@

@@

% Comparison Markov class standard PC and EMS-PC
Both standard PC and EMS-PC find many bi-directional arrows, with some one-directional arrows.
As can be seen in the distribution of values in figure~\ref{fig:distributions}, nearly all found edges are oriented as bi-directional arrows.
For EMS-PC, this number is somewhat lower but there are still few one-directional arrows, and they do not have certainties higher than 0.15.
These certainties are not high enough to conclude any causal relations.

@@

@@

% TODO: Suggereren task based analysis ipv resting state voor beter causaal beeld.

%cause of poor consistency: inherent to our method or viable for further research?

% \section{Conclusion}
%repeat everything in 1 paragraph, or leave away, whatever seems most fitting in the end. HRF makes us awesome

>>>>

% Margins & Headers
%\marginsize{2.5cm}{2.5cm}{3.0cm}{2.0cm}
%\columnsep 0.4in
%\footskip 0.4in 
\usepackage{changepage}

@@

\paragraph{Run-time performance}
The PC algorithm run-time is in the order of the amount of dependency tests to be performed and thus it has a worst-case runtime complexity exponential in the maximal branching degree present in the graph.
However, this worst-case scenario only happens when no edges can be removed; all subsets of the complete set of nodes have to be used to test \textit{d}-separation of every pair of nodes.
For a sparse graph, many edges may be removed for small separating sets, and this enables the standard PC algorithm to often run fast in practice perform well on hundreds of variables.

@@

%EMS-PC extension
\subsection{EMS-PC}
We have combined the above ideas - multiple separating sets, unfaithfulness check and explicit orientation check - into an adaptation of PC, which we refer to as Explicit Multiple Separating Set PC (EMS-PC).
These three changes have been made in this adaption with respect to standard PC as described in algorithm~\ref{alg:pc_alg}.
At line~\ref{alg:pc_alg:next_pair}, the search for separating sets for $X$ and $Y$ stops when a separating set is found; instead, EMS continues the search for more separating sets of the same order.
All of these separating sets are recorded.

@@

\subsection{Experimental validation}
% What type of data is needed
PC algorithm requires observational data where each instance is a vector of single observations per variable.
We can model functional resting-state time-series as coming from some stable underlying mechanism that is largely time-independent.
Resting-state data can be intuitively thought of being more time independent than task-influenced data due to the emergent character of activity pattens that occur when in resting-state.
Nevertheless it is most likely not a fully correct modelling assumption, though enables straightforward application of standard and EMS-PC to functional data without the problem of HRF.

\paragraph{Data acquisition}
We applied standard PC and EMS-PC on functional resting-state time-series data of six healthy individuals, also previously used by Hinne et al.\cite{hinne2013}.
After preprocessing the fMRI data was parcellated according to the Automated Anatomical Labeling (AAL) atlas in 116 regions of interest.
This provides 1029 samples with 116 standard-normally distributed variables.
For the exact procedure of subject acquisition, used scanner properties and parameters and data preprocessing we refer to the paper by Hinne et al. 
Additionally, diffusion-weighted images of these subjects have been obtained from the same study.
These were provided as six $116 \times 116$ binary connectivity matrices.
No additional preprocessing was needed for application of our methods.
We use Fisher's z-test for conditional independence to match the continuous nature of the data.

\paragraph{Implementation}
We have implemented standard PC and EMS-PC in Matlab using statistical tests and ideas from the Bayesian Network Toolbox for Matlab.\footnote{Bayesian Network Toolbox for Matlab: https://github.com/bayesnet/bnt}
This toolbox also includes an implementation for standard PC, however to gain better insight in the workings and to have greater flexibility in application, we have implemented standard PC separately.
The Matlab scripts used for the results are made freely available online.\footnote{Matlab implementation of EMS-PC: https://github.com/tomderuijter/EMS-PC}

@@

\paragraph{Parameter selection}
Due to the lack of a baseline to compare results with, dependency thresholds were chosen by hand.
All experiments listed here were run with an alpha value of 0.01 for the independence test.
The number of random permutations was set on 1000 per subject, which was found to be enough to determine statistical significance for 99\% of the found edges with a binomial statistical test.

@@

\section{Results}\label{sec:results}
Due to the size of the resulting graphs, we represent them in the form of connectivity matrices for convenience.
A connection from region $n$ to region $m$ is represented by point $(n,m)$ in the matrix.
Higher values represent higher frequencies of a connection over all permutations.
In all matrices, indices 1-45 represent the brain regions of the left hemisphere, indices 45-90 represent the right hemisphere, and indices 91-116 represent regions in the cerebellum.
As such, the matrices are divided into blocks: the top left corner represents the connections within the left hemisphere, the bottom left corner represents the connections between the left hemisphere and the cerebellum, and so on.

\subsection{Structural graphs}
% Measuring whole-brain structural and directional consistency of PC - frequency of occurrences within subjects
% Applying the standard PC algorithm yields an undirected graph, in which each edge has a certainty.
% This is done both per subject, and as an avarage over all subjects.
% We compared the generated structure with a Bayesian inference approach on finding structural connectivity from diffusion weighted imaging, as described in Hinne et al.\cite{hinne2013}.
Modified PC and EMS-PC do not differ in finding functional connectivity.
As edges found with the structural part of PC are undirected, the found matrices are symmetrical.
The results are presented per subject in figure~\ref{fig:struct_subjects} and as an average across all subjects in figure~\ref{fig:struct_avg}.
For comparison, the structural connectivity as found by Hinne et al.\cite{hinne2013} are also show.

@@

\subsection{Directed graphs}
This structural graph has been used as a basis for a directed graph, for both standard PC and EMS-PC.
As there were little to none notable differences between different subjects, only the average DAG across all subjects is presented.
The results for standard PC and EMS-PC are shown in figure~\ref{fig:pdag_avg}.
Although many arrowheads are found with frequency ratio above 0.9 for both variants of the algorithm, the graph looks strongly symmetric.
In terms of causality, this means that many latent sources are still present and causal relations between the measured brain regions themselves are not clearly visible.

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.49\textwidth]{images/new/PDAG_avg_mod-crop}
  \includegraphics[width=0.49\textwidth]{images/new/PDAG_avg_expl-crop}
  \caption{The partially DAG found by standard PC on the left, and EMS-PC on the right. A high value denotes a high certainty of an arrowhead. The sum of edge frequencies of all six subjects is taken.}
  \label{fig:pdag_avg}
\end{figure}

Another representation of this data has been used to emphasise the asymmetric parts of the directed graphs.
To achieve this, the absolute difference between these matrices and their transpose has been calculated, lifting out those points for which there is an arrow in one direction, but not in the opposite direction.
The resulting antisymmetric matrices are presented per subject for EMS-PC in figure~\ref{fig:asym_subjects}. The average across all subjects for both variants of the PC algorithm are presented in figure~\ref{fig:asym_avg}.

\begin{figure}[h!]
  \centering
  \includegraphics[height=0.25\textwidth]{images/new/assym_subj1_expl-crop}
  \includegraphics[height=0.25\textwidth]{images/new/assym_subj2_expl-crop}
  \includegraphics[height=0.25\textwidth]{images/new/assym_subj3_expl-crop}
  \includegraphics[height=0.25\textwidth]{images/new/assym_subj4_expl-crop}
  \includegraphics[height=0.25\textwidth]{images/new/assym_subj5_expl-crop}
  \includegraphics[height=0.25\textwidth]{images/new/assym_subj6_expl-crop}
  \caption{The asymetric parts of the partial DAG found with EMS-PC for all six subjects, in which a high value denotes a high certainty of a one-directional arrow.}
\label{fig:asym_subjects}
\end{figure}

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.49\textwidth]{images/new/assym_avg_mod-crop}
  \includegraphics[width=0.49\textwidth]{images/new/assym_avg_expl-crop}
  \caption{The asymmetric parts of the partial DAG found with standard PC on the left, and EMS-PC on the right. A high value denotes a high certainty of an one-directional arrow. The average of all six subjects is taken.}
  \label{fig:asym_avg}
\end{figure}

@@

Our additional unfaithfulness check without additional independence tests were unable to detect any unfaithful v-structures.

@@

% Comparison Markov class standard PC and EMS-PC
Both standard PC and EMS-PC find many bi-directional arrows, with some one-directional arrows.
As can be seen in the distribution of values in figure~\ref{fig:distributions}, nearly all found edges are oriented as bi-directional arrows.
For EMS-PC, this number is somewhat lower but there are still few one-directional arrows, and they do not have certainties higher than 0.15.
These certainties are not high enough to conclude any causal relations.
The additional unfaithfulness test proved inconclusive, as no unfaithful v-structures were found.
This is not bad news, but it does suggest there are methodological details being skipped over.

@@

Instead of targeting the brain as a whole in resting-state, one could focus more on specific parts to increase the modelling detail.
In such a case task-based data would work well.
An area corresponding to a specific task, for instance the motor cortex, could be inspected in relationship with itself and other areas.

@@
